The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R

### 2023-11-15 manuscript/manuscript.qmd
implemented reviewer #1 comments/revisions

[Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi>. Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.

@@

While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python. Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community. As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.

@@

@@

Despite the potential initial complexities to get acquainted with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why. Git makes it easier for users to go back to any snapshot and resolve errors more efficiently. Further, remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code. Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020] and PCORnet[@fleurence2014], which makes shared code interoperable even without direct access to the same data source. Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.

>>>>

[Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2. All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi> or <https://github.com/janickweberpals/fair-epi>. Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.

@@

While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking. Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community. As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.

@@

## Creating persistent identifiers

A cornerstone of FAIR principles is the use of persistent identifiers, such as digital object identifiers (DOI), for enabling findability and accessibility of research objects. Integrating the concept of persistent identifiers within the context of remote repositories is an important aspect to enhance the traceability and accessibility of code repositories or even specific commits or releases of a study codebase.

To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019;@zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts. The platform seamlessly interfaces with GitHub, enabling the creation of referenceable DOIs for existing code repositories. The steps required to obtain a DOI involve the linkage of a Zenodo and a GitHub account[@zenodoGitHub]. Once linked, a user will be able to select a specific release of a repository which automatically registers a new DOI and a corresponding badge which can be included and referenced in a GitHub README file.

@@

Despite the potential initial complexities to get acquainted with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why. Remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code. While the access to the final version of the programming code is already sufficient for external reproducibility, the tracking of changes over time enables a study team to comprehend and resolve errors more efficiently. Similar to tracking changes and amendments of a study protocol, an external audience also benefits from a complete audit trail through a better understanding of the documented scientific rationale behind specific changes, the identification of implemented quality control processes (e.g., code reviews) and the tracking of contributions of different collaborators to the codebase.

Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020], PCORnet[@fleurence2014] and ConcePTION [@thurin2021], which makes shared code interoperable even without direct access to the same data source. In order to achieve full replicability, authors may alternatively include a sample of dummy data to enable readers to comprehend and run the analytic code. Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).\[@gonzal

Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.
